---
title: "Estimating tree biomass and merchantable volume"
author: "Hunter Stanke, Jeffrey W. Doser"
date: "2021 (last updated February 7, 2025)"
description: Some basics on estimating tree biomass and volume in `rFIA` 
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{biomassVolume}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

Prior to `rFIA` v0.3.1, the `biomass()` function provided estimates of volume, biomass, and carbon stocks of standing trees. This is what is described in the introductory `rFIA` manuscript ([Stanke et al. 2020](https://research.fs.usda.gov/treesearch/59521).

`rFIA` v0.3.1 introduced breaking changes to the `biomass()` function and how `rFIA` handles volume, carbon, and biomass estimation. Specifically, the `biomass()` function was overhauled to focus solely on estimation of tree biomass, providing enhanced flexibility in the estimation of *biomass components* (e.g., bole, limbs, coarse roots, etc.). In addition, v0.3.1 gave rise to the `volume()` function, which supersedes and expands upon the previous merchantable volume estimators implemented in `biomass()`. Carbon estimation now lives entirely in the dedicated `carbon()` function, which estimates carbon stocks across the full suite of forest ecosystem carbon pools (e.g., live and dead trees, understory vegetation, down dead wood, litter, soil organic matter), not just standing trees; see `?carbon` for details.

As of `rFIA` v1.1.0, the `biomass()` and `volume()` functions now leverage FIA's new National Scale Volume and Biomass Estimators (NSVB), which provides a more consistent and accurate accounting of structural components of trees. As with previous versions of `biomass()`, estimates can be returned by tree component using the `byComponent` method, although we note the specific components available have changged given underlying changes in FIADB and the NSVB approach. The NSVB represents a substantial advancement in national scale volume and biomass equations. Below we provide some basic examples of `biomass()` and `volume()`.

# Estimating tree biomass components

Let's see how the `biomass()` function can be used to estimate tree biomass components! As always, you can check out our documentation with `?biomass`:

```{r, echo=TRUE, eval=TRUE, warning=FALSE}
# Load data from the rFIA package
library(rFIA)
data(fiaRI)

# Total live tree aboveground biomass, excluding foliage (default)
biomass(fiaRI)
```

By default, `biomass` estimates total aboveground live tree biomass for all reporting years available in the `FIA.Database` (`treeType = "live"`, `component = "AG"`). Here aboveground tree biomass is a simple summation of wood biomass in the total stem from ground line to the tree tip, stem bark biomass, and branch wood and bark.

But what if we're interested in estimating biomass within different tree components separately, e.g., we want an individual estimate for bole, foliage, stump, and others? Simply set the new `byComponent` argument to `TRUE`:

``` {r, echo=TRUE, eval=TRUE, warning=FALSE}
biomass(db = clipFIA(fiaRI), byComponent = TRUE)
```

Note that the sum of the provided components is NOT equal to total tree biomass. In other words, the components are not mutually exclusive (e.g., `BOLE` biomass is part of `STEM` biomass). Aboveground tree biomass is equal to the sum of `STEM`, `STEM_BARK`, and `BRANCH`. The *Estimation Details* section of the help page for `biomass()` provides all this information and more.  

Awesome, but what if we want to estimate biomass for some combination of these components, e.g., bole plus stump? The new `component` argument has our backs. Users can specify any combination of components seen in the output above. For example, say we want to estimate abovegound biomass (`"AG"`) plus foliage (`"FOLIAGE"`):

``` {r, echo=TRUE, eval=TRUE, warning=FALSE}
biomass(db = clipFIA(fiaRI), component = c("AG", "FOLIAGE"))
```

# Estimating merchantable tree volume

Previously, `biomass()` included support for estimation of net merchantable volume and net sawlog volume in units of cubic feet. Since `rFIA` v0.3.1, the `volume()` function expands on this previous capacity in two key ways: (1) allowing use of alternative volume definitions used by the FIA program (i.e., net, sound, and gross volume), and (2) offering estimates of sawlog volume in units of cubic feet (CF) and thousand board feet (MBF; International 1/4 inch rule). 

By default, `volume()` will estimate net volume of live trees (`volType = "NET"` and `treeType = "live"`):

``` {r echo=TRUE, eval=TRUE, warning=FALSE}
volume(db = fiaRI)
```

Here, `BOLE_CF_ACRE` gives us merchantable bole volume per acre, `SAW_CF_ACRE` gives us sawlog volume in cubic feet per acre, and `SAW_MBF_ACRE` gives us sawlog volume in thousand board feet per acre (International 1/4 inch rule). 

We can change our volume definition using the `volType` argument. Let's try gross volume instead:
``` {r echo=TRUE, eval=TRUE, warning=FALSE}
volume(db = fiaRI, volType = 'gross')
```

So what do these different definitions mean? FIA defines net volume (`volType="NET"`) as: "The net volume of wood in the central stem of timber species (trees where diameter is measured at breast height $\geq$ 5.0 inches d.b.h., from a 1-foot stump to a minimum 4-inch top diameter, or to where the central stem breaks into limbs all of which are <4.0 inches in diameter.... Does not include rotten, missing, and form cull (volume loss due to rotten, missing, and form cull defect has been deducted)." In `volum()`, we could also choose from two alternative definitions: sound volume (`volType = 'SOUND'`) or gross volume (`volType = 'GROSS'`). Sound volume is identical to net volume except that sound includes volume from portions of the stem that are be considered "form cull" under the net volume definition (e.g., sweep). In contrast, gross volume is identical to the net volume definition except that gross includes volume from portions of the stem that are rotten, missing, and considered form cull.

