---
title: "Validating rFIA against EVALIDator"
author: "Jeffrey W. Doser"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
description: A running record of unit tests that compare `rFIA` estimates to EVALIDator
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{evalidatorValidation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(knitr.kable.NA = "")

# Directory holding the rFIA test logs. Each file is the full output of one
# run of the test suite (rfia-tests-YYYY-MM-DD_HHMM.txt). Set the
# RFIA_TEST_LOGS environment variable to render from a different directory.
logDir <- Sys.getenv("RFIA_TEST_LOGS", "~/rfia-test-logs")
logFiles <- sort(list.files(logDir, pattern = "^rfia-tests-.*\\.txt$", full.names = TRUE))
haveLogs <- length(logFiles) > 0
```

```{r parse, include = FALSE, eval = haveLogs}
# Pulls the value of a "Key:   value" line from the log header.
headerValue <- function(lines, key) {
  hit <- grep(paste0("^", key, ":"), lines, value = TRUE)
  if (length(hit) == 0) return(NA_character_)
  trimws(sub(paste0("^", key, ":"), "", hit[1]))
}

# Parses a single test log into run-level metadata and (when the log
# includes one) the per-test results table.
parseLog <- function(path) {
  lines <- readLines(path, warn = FALSE, encoding = "UTF-8")
  started <- headerValue(lines, "Run started")
  totals <- grep("^\\[ FAIL [0-9]+ \\|", lines, value = TRUE)
  getTotal <- function(what) {
    if (length(totals) == 0) return(NA_integer_)
    as.integer(sub(paste0(".*", what, " ([0-9]+).*"), "\\1", totals[1]))
  }
  overall <- headerValue(lines, "Overall")
  meta <- data.frame(
    log = basename(path),
    started = as.POSIXct(sub(" [A-Z]+$", "", started), format = "%Y-%m-%d %H:%M:%S"),
    commit = headerValue(lines, "Git commit"),
    tree = headerValue(lines, "Working tree"),
    filter = headerValue(lines, "Test filter"),
    suite = if (is.na(overall)) "Runner error" else if (grepl("PASSED", overall)) "Passed" else "Failed",
    suitePass = getTotal("PASS"),
    suiteFail = getTotal("FAIL"),
    suiteSkip = getTotal("SKIP"),
    stringsAsFactors = FALSE
  )
  start <- grep("^==== PER-TEST RESULTS \\(CSV\\) ====$", lines)
  end <- grep("^==== END PER-TEST RESULTS ====$", lines)
  tests <- NULL
  if (length(start) == 1 && length(end) == 1 && end > start + 1) {
    tests <- read.csv(text = lines[(start + 1):(end - 1)], stringsAsFactors = FALSE)
    tests$log <- basename(path)
  }
  list(meta = meta, tests = tests)
}

parsed <- lapply(logFiles, parseLog)
runs <- do.call(rbind, lapply(parsed, `[[`, "meta"))
tests <- do.call(rbind, lapply(parsed, `[[`, "tests"))

# EVALIDator comparisons are identified by test name: every test that checks
# an rFIA estimate against an EVALIDator reference value contains
# "EVALIDator" in its description (e.g., "tpa() matches EVALIDator for ...").
# The function is read off the start of the test name, falling back to the
# test file name.
if (!is.null(tests)) {
  evTests <- tests[grepl("EVALIDator", tests$test), ]
  evTests$fn <- ifelse(grepl("^[A-Za-z]+\\(\\)", evTests$test),
                       sub("^([A-Za-z]+)\\(\\).*", "\\1", evTests$test),
                       sub("^test-(.*)\\.R$", "\\1", evTests$file))
  evTests$status <- ifelse(evTests$failed > 0 | evTests$error, "Failed",
                           ifelse(evTests$skipped, "Skipped", "Passed"))
} else {
  evTests <- NULL
}

runs$hasDetail <- runs$log %in% unique(tests$log)
fmtDate <- function(x) format(x, "%Y-%m-%d %H:%M")

# Latest run with per-test detail, preferring runs of the full suite (no
# test filter).
detailRuns <- runs[runs$hasDetail, ]
fullRuns <- detailRuns[is.na(detailRuns$filter), ]
if (nrow(fullRuns) > 0) detailRuns <- fullRuns
latest <- if (nrow(detailRuns) > 0) detailRuns[which.max(detailRuns$started), ] else NULL
latestEv <- if (!is.null(latest)) evTests[evTests$log == latest$log, ] else NULL
```

```{r nologs, results = 'asis', eval = !haveLogs}
cat("> **Note:** no test logs were found in `", logDir, "`, so the results ",
    "tables below could not be generated. Set the `RFIA_TEST_LOGS` ",
    "environment variable to the directory holding the logs and re-render.\n", sep = "")
```

# Overview

EVALIDator is the USDA Forest Service's online tool for producing
population estimates from the FIA Database (FIADB). Its estimates are the
reference that FIA analysts and users compare against, so `rFIA` must
reproduce them. The `rFIA` unit tests check estimates from the package's main
estimation functions against values from EVALIDator for a range of common
estimation settings. This document summarizes the results of those tests and
keeps a record of how the tests have performed over time.

## How the comparisons work

Each comparison test has the same structure:

1. **Data.** We read FIADB data for four states, one from each FIA region:
   Rhode Island (Northern), North Carolina (Southern), Colorado (Interior West),
   and Oregon (Pacific Northwest). We then subset each state to its most recent
   evaluation with `clipFIA(mostRecent = TRUE)`.
2. **Reference values.** At test time, reference values come from the
   FIADB-API `fullreport` endpoint, the programmatic interface behind
   EVALIDator. We request the same evaluation group (`EVAL_GRP`), attribute
   (e.g., aboveground biomass of live trees on forest land), and domain filter
   (e.g., a physiographic class filter for `areaDomain`, or a species filter
   for `treeDomain`) that the `rFIA` call uses.
3. **Comparison.** `rFIA` is run with the temporally indifferent estimator
   (the default `method = 'TI'`), which is the method EVALIDator uses. Point
   estimates and percent sampling errors must match EVALIDator to within a
   relative tolerance of 10<sup>-6</sup>. Where EVALIDator reports them, the
   number of non-zero plots used in the numerator and denominator of ratio
   estimates must match exactly.

The settings tested include the default case for each function, `landType`
and `treeType` variants, biomass and carbon components, `treeDomain` and
`areaDomain` filters, and per-species estimates (`bySpecies = TRUE`). One test
is an approximate comparison rather than an exact one: `growMort()` for
Oregon timberland. It checks that a known, documented discrepancy stays below
1% (see [Known EVALIDator discrepancies](#known-evalidator-discrepancies)).
The full test code is in `tests/testthat/` of the
[rFIA GitHub repository](https://github.com/doserjef/rFIA). The analysis
behind each function's tests is described in `core_references/validation/`.

A test is recorded as **skipped**, not failed, when the FIADB-API cannot be
reached. A skip means the comparison did not run, not that `rFIA` disagreed
with EVALIDator.

## How this document is updated

The full `rFIA` test suite runs on a regular schedule (every 2 weeks) in addition
to automated testing via GitHub. This document is updated each time a new log file
is generated.

```{r, results = 'asis', eval = haveLogs}
cat("This version was generated from ", nrow(runs), " test log(s) found in `",
    logDir, "`, covering runs from ", fmtDate(min(runs$started)), " to ",
    fmtDate(max(runs$started)), ".\n", sep = "")
```

# Known EVALIDator discrepancies

In one case, `rFIA` deliberately does not reproduce EVALIDator. We describe it
below, along with why we chose not to match EVALIDator.

## Timberland growth and mortality in states that use macroplots

**Affected functions:** `growMort()` and `vitalRates()` with
`landType = 'timber'`.

**Affected states:** states whose condition proportions are measured on the
macroplot, mainly in the Pacific Northwest (e.g., Oregon, Washington, and
California). Estimates for forest land (`landType = 'forest'`, the default)
match EVALIDator exactly in every state. Timberland estimates match
EVALIDator exactly in Rhode Island, North Carolina, and Colorado, which have
no macroplot-based conditions.

**Size of the discrepancy:** at the time of the investigation (August 2026,
Oregon evaluation group 412203), `rFIA` counted 7,788 plots in the Oregon
timberland area denominator, compared with 7,749 in EVALIDator. This is a
difference of 39 plots (about 0.5%). The resulting per-acre estimates differ
by about 0.09%. For example, annual mortality was 1.672704 trees per acre in
`rFIA` and 1.671130 in EVALIDator, and aboveground biomass growth was 1.367651
in `rFIA` and 1.366364 in EVALIDator. Plot counts differed by a similar
margin in California (2,580 vs. 2,562) and Washington (4,899 vs. 4,870).

### What causes the difference

Growth, removal, and mortality estimates are ratios. The denominator is the
area of land that was in the same land class (forest land or timberland) at
both the previous and current measurements of a plot. FIA estimates this area
from the `SUBP_COND_CHNG_MTRX` table, which records the proportion of each
subplot or macroplot that moved between condition classes between the two
measurements. Each condition's area proportion is measured on either the
subplot (`COND.PROP_BASIS = 'SUBP'`) or the larger macroplot
(`PROP_BASIS = 'MACR'`). The FIA Population Estimation User Guide (Chapter
7.8) says to take the change proportions from the matching plot component:
`SUBPTYP = 1` rows for subplot-based conditions and `SUBPTYP = 3` rows for
macroplot-based conditions.

`rFIA` applies this rule to both forest land and timberland. We obtained the
SQL that EVALIDator runs for these estimates (the FIADB-API returns it with
each estimate) and found that EVALIDator is not consistent between the two
land classes:

- For the **forest land** denominator, EVALIDator applies the same
  subplot/macroplot rule as `rFIA`, and the two agree exactly.
- For the **timberland** denominator, EVALIDator only reads the subplot rows
  (`SUBPTYP = 1`), so change on macroplot-based conditions is never read.
  It then multiplies these subplot proportions by the *macroplot* adjustment
  factor (`ADJ_FACTOR_MACR`) for macroplot-based conditions. A subplot
  proportion scaled by a macroplot adjustment factor does not correspond to a
  meaningful area.

### Why `rFIA` does not reproduce EVALIDator here

We decided to keep the `rFIA` behavior for the following reasons: 

1. **It follows the documented method.** `rFIA` implements the rule from the
   FIA Population Estimation User Guide. We found no documentation that gives
   a reason for leaving macroplot-based timberland out of the
   growth-accounting area, and our own assessment of the database logic is in 
   line with the `rFIA` implementation.
2. **It is consistent across land classes.** EVALIDator uses the
   subplot/macroplot rule for forest land but not for timberland. Reproducing it 
   would make `rFIA` treat the two land classes differently for no reason beyond
   simply matching EVALIDator.
3. **The effect is small and limited.** The difference is under 1% and only
   occurs for timberland in states that use macroplot-based conditions.

### How the tests handle it

The `growMort()` tests compare Oregon timberland mortality to EVALIDator
approximately: they check that the relative difference stays below 1%, so the
test will catch any growth in the discrepancy. The `vitalRates()` timberland
comparisons run only for Rhode Island, North Carolina, and Colorado, where the
two agree exactly. 

# Latest results

```{r latestHeader, results = 'asis', eval = haveLogs}
if (is.null(latest)) {
  cat("None of the available logs include per-test results, so the number of",
      "EVALIDator comparisons cannot be determined yet. Per-test results are",
      "recorded in logs from runs made after this document was added.\n")
} else {
  cat("Results below are from the test run that started on **", fmtDate(latest$started),
      "** (commit `", latest$commit, "`",
      if (!is.na(latest$tree) && latest$tree != "clean") paste0(", with ", latest$tree) else "",
      ").\n", sep = "")
}
```

## Summary by function

```{r latestTable, eval = haveLogs && !is.null(latest)}
summarizeFn <- function(d) {
  fns <- sort(unique(d$fn))
  out <- data.frame(
    Function = paste0("`", fns, "()`"),
    Tests = as.integer(table(d$fn)[fns]),
    Checks = as.integer(tapply(d$nb, d$fn, sum)[fns]),
    Passed = as.integer(tapply(d$status == "Passed", d$fn, sum)[fns]),
    Failed = as.integer(tapply(d$status == "Failed", d$fn, sum)[fns]),
    Skipped = as.integer(tapply(d$status == "Skipped", d$fn, sum)[fns]),
    stringsAsFactors = FALSE
  )
  out <- rbind(out, data.frame(Function = "**Total**", Tests = sum(out$Tests),
                               Checks = sum(out$Checks), Passed = sum(out$Passed),
                               Failed = sum(out$Failed), Skipped = sum(out$Skipped)))
  out$Status <- ifelse(out$Failed > 0, "Failing",
                       ifelse(out$Passed == 0, "Not run (skipped)",
                              ifelse(out$Skipped > 0, "Passing (some skipped)", "Passing")))
  out
}
knitr::kable(summarizeFn(latestEv), align = "lrrrrrl",
             caption = paste("EVALIDator comparison tests in the run of",
                             fmtDate(latest$started), "by function."))
```

`Tests` is the number of EVALIDator comparison tests for each function; most
settings are tested once in each of the four states. `Checks` is the total
number of individual assertions in those tests (point estimates, percent
sampling errors, and plot counts). A test passes only if all of its checks
pass.

## Failed and skipped comparisons

```{r latestProblems, results = 'asis', eval = haveLogs && !is.null(latest)}
bad <- latestEv[latestEv$status != "Passed", ]
if (nrow(bad) == 0) {
  cat("All EVALIDator comparison tests passed in the latest run.\n")
} else {
  print(knitr::kable(data.frame(Function = paste0("`", bad$fn, "()`"),
                                Test = bad$test, Status = bad$status),
                     row.names = FALSE))
}
```

# History across test runs

## Summary by run

```{r history, eval = haveLogs}
hist <- runs[order(runs$started, decreasing = TRUE), ]
evCount <- function(lg, what) {
  if (!lg %in% unique(evTests$log)) return(NA_integer_)
  d <- evTests[evTests$log == lg, ]
  if (what == "Tests") nrow(d) else sum(d$status == what)
}
histTab <- data.frame(
  Run = fmtDate(hist$started),
  Commit = paste0("`", sub(" .*", "", hist$commit), "`",
                  ifelse(!is.na(hist$tree) & hist$tree != "clean", "*", "")),
  Tests = sapply(hist$log, evCount, what = "Tests"),
  Passed = sapply(hist$log, evCount, what = "Passed"),
  Failed = sapply(hist$log, evCount, what = "Failed"),
  Skipped = sapply(hist$log, evCount, what = "Skipped"),
  Suite = ifelse(is.na(hist$filter), hist$suite, paste0(hist$suite, " (subset)")),
  stringsAsFactors = FALSE
)
histTab$Suite <- ifelse(is.na(hist$suitePass), histTab$Suite,
                        paste0(histTab$Suite, " (", hist$suitePass, " passed, ",
                               hist$suiteFail, " failed, ", hist$suiteSkip, " skipped)"))
knitr::kable(histTab, row.names = FALSE, align = "llrrrrl",
             caption = "EVALIDator comparison results for each test run, newest first.")
```

`Tests`, `Passed`, `Failed`, and `Skipped` count EVALIDator comparison tests.
They are blank for older logs, which did not record per-test results. `Suite`
gives the overall result of the full `rFIA` test suite, including tests that
do not involve EVALIDator, with counts of individual checks. An asterisk next
to a commit means the run was made with uncommitted changes to the package
source. "(subset)" marks runs made with only part of the test suite.

## Summary by function and run

```{r historyByFn, eval = haveLogs && !is.null(evTests) && nrow(evTests) > 0}
detailLogs <- hist$log[hist$log %in% unique(evTests$log)]
fns <- sort(unique(evTests$fn))
byFn <- sapply(detailLogs, function(lg) {
  d <- evTests[evTests$log == lg, ]
  sapply(fns, function(f) {
    df <- d[d$fn == f, ]
    if (nrow(df) == 0) return("--")
    paste0(sum(df$status == "Passed"), "/", nrow(df))
  })
})
byFn <- matrix(byFn, nrow = length(fns), dimnames = list(NULL, NULL))
colnames(byFn) <- fmtDate(hist$started[match(detailLogs, hist$log)])
knitr::kable(data.frame(Function = paste0("`", fns, "()`"), byFn, check.names = FALSE),
             row.names = FALSE,
             caption = "Passed / total EVALIDator comparison tests by function, newest run first.")
```

# Appendix: all EVALIDator comparisons in the latest run

```{r appendix, eval = haveLogs && !is.null(latest)}
app <- latestEv[order(latestEv$fn, latestEv$test), ]
knitr::kable(data.frame(Function = paste0("`", app$fn, "()`"), Test = app$test,
                        Checks = app$nb, Status = app$status),
             row.names = FALSE, align = "llrl")
```
