---
title: "Estimating individual-tree and population growth rates"
author: "Hunter Stanke, Jeffrey W. Doser"
date: "2021 (last updated February 7, 2025)"
description: Basic tools for estimating growth with `rFIA`
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{growth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

We can use FIA data to estimate how forest structure has changed over time (i.e., has density increased or declined) or to estimate the average rate at which individual trees grow. Here we show how we can use the `vitalRates()` function in `rFIA` to do just that.

For this specific example, we'll focus on estimating average annual change in individual tree basal area and cumulative tree basal area per acre, by site productivity class (SITECLCD) and species across the states of Washington and Oregon. 

So first, we'll need to download some data (you can skip this step if you already have FIA data for WA and OR). Here we use `getFIA()` to download and save the state subsets to our computer:

```{r, eval=FALSE, echo=TRUE}
library(rFIA)
library(dplyr)

# Download some FIA data from the DataMart ------------------------------------
getFIA(states = c('OR', 'WA'),
       dir = 'path/to/save/FIA/data',
       load = FALSE) # Download, but don't load yet
```

```{r, eval = TRUE, message = FALSE, warning = FALSE, echo = FALSE}
library(rFIA)
library(dplyr)
```

Next we'll read our data into R with `readFIA()` and take a most recent subset across the two states with `clipFIA()`. Note that we are setting `inMemory = FALSE` to set up the database in a remote fashion as opposed to reading the entire data set into RAM. You would need to set `inMemory = TRUE` if you desire to modify columns in the database or do any other sort of manipualation to the underlying data.

```{r, eval = FALSE, echo = TRUE}
db <- readFIA(dir = fiaPath,
              states = c('OR', 'WA'), # If you keep all your data together
              nCores = 1, # Can set to >1 if you have multiple cores.
              inMemory = FALSE) # Set to TRUE if your computer has enough RAM

# Take the most recent subset -------------------------------------------------
db <- clipFIA(db, mostRecent = TRUE)
```

```{r, eval = TRUE, echo = FALSE}
db <- readFIA(dir = '~/Dropbox/data/fia/',
              states = c('OR', 'WA'), # If you keep all your data together
              nCores = 1, # Can set to >1 if you have multiple cores.
              inMemory = FALSE) # Set to TRUE if your computer has enough RAM

# Take the most recent subset -------------------------------------------------
db <- clipFIA(db, mostRecent = TRUE)
```

Now that we have our data loaded, we can start estimating tree growth rates! First, we'll estimate *net* growth rates, i.e., we will include trees that have recruited or died in our estimates of tree growth. Note that net growth can be negative if mortality exceeds recruitment and growth on live trees. This type of growth estimate is most useful if we're interested in characterizing population-level change. For example, if we'd like to know how the average rate at which cumulative basal area per acre has changed in our population of interest, we'd shoot for net growth. Here's how we do it with `rFIA`, where net growth is indicated by the argument `treeType = 'all'`:

```{r, eval = TRUE, echo=TRUE}
# Net annual change
net <- vitalRates(db,
                  treeType = 'all', # "all" indicates net growth
                  grpBy = SITECLCD, # Grouping by site productivity class
                  bySpecies = TRUE) # also grouping by species
net
```

If we are instead interested in characterizing the average annual growth rate of individual trees, we'd most likely want to exclude stems that died or recruited into the population between plot measurements. To do this with rFIA, simply set `treeType = 'live'`:

```{r, eval=TRUE, echo=TRUE}
# Annual growth on live trees that remained live
live <- vitalRates(db,
                   treeType = 'live', # "live" excludes mortality and recruitment
                   grpBy = SITECLCD, # Grouping by site productivity class
                   bySpecies = TRUE) # also grouping by species
```

By default, `vitalRates()` will estimate average annual DBH, basal area, biomass, and net volume growth rates for *individual stems*, along with average annual basal area, biomass, and net volume growth *per acre*. Here we'll focus in on basal area, so let's simplify and pretty up our tables a bit:

```{r, eval=TRUE, echo=TRUE}
# Net annual change
net <- net %>%
  # Dropping unnecessary columns
  select(SITECLCD, COMMON_NAME, SCIENTIFIC_NAME, SPCD,
         BA_GROW, BA_GROW_AC, BA_GROW_SE, BA_GROW_AC_SE, nPlots_TREE) %>%
  # Dropping rows with no live trees (growth was NA)
  filter(!is.na(BA_GROW)) %>%
  # Making SITECLCD more informative
  mutate(site = case_when(SITECLCD == 1 ~ "225+ cubic feet/acre/year",
                          SITECLCD == 2 ~ "165-224 cubic feet/acre/year",
                          SITECLCD == 3 ~ "120-164 cubic feet/acre/year",
                          SITECLCD == 4 ~ "85-119 cubic feet/acre/year",
                          SITECLCD == 5 ~ "50-84 cubic feet/acre/year",
                          SITECLCD == 6 ~ "20-49 cubic feet/acre/year",
                          SITECLCD == 7 ~ "0-19 cubic feet/acre/year")) %>%
  # Arrange it nicely
  select(COMMON_NAME, SCIENTIFIC_NAME, SITECLCD, site, everything()) %>%
  arrange(COMMON_NAME, SCIENTIFIC_NAME, SITECLCD, site)

# Annual growth on live trees that remained live
live <- live %>%
  # Dropping unnecessary columns
  select(SITECLCD, COMMON_NAME, SCIENTIFIC_NAME, SPCD,
         BA_GROW, BA_GROW_AC, BA_GROW_SE, BA_GROW_AC_SE, nPlots_TREE) %>%
  # Dropping rows with no live trees (growth was NA)
  filter(!is.na(BA_GROW)) %>%
  # Making SITECLCD more informative
  mutate(siteProd = case_when(SITECLCD == 1 ~ "225+ cubic feet/acre/year",
                              SITECLCD == 2 ~ "165-224 cubic feet/acre/year",
                              SITECLCD == 3 ~ "120-164 cubic feet/acre/year",
                              SITECLCD == 4 ~ "85-119 cubic feet/acre/year",
                              SITECLCD == 5 ~ "50-84 cubic feet/acre/year",
                              SITECLCD == 6 ~ "20-49 cubic feet/acre/year",
                              SITECLCD == 7 ~ "0-19 cubic feet/acre/year")) %>%
  # Arrange it nicely
  select(COMMON_NAME, SCIENTIFIC_NAME, SITECLCD, siteProd, everything()) %>%
  arrange(COMMON_NAME, SCIENTIFIC_NAME, SITECLCD, siteProd)
```

Here `BA_GROW` gives us annual basal area growth per tree in square feet/year, and `BA_GROW_AC` gives us average basal area growth per acre in square feet/acre/year. But maybe we'd prefer units to be square centimeters instead.

```{r, eval = TRUE, echo=TRUE}
# Net annual change
net <- net %>%
  # Convert to square centimeters instead of square feet
  mutate(BA_GROW = BA_GROW * 929.03,
         BA_GROW_AC = BA_GROW_AC * 929.03)
net

# Annual growth on live trees that remained live
live <- live %>%
  # Convert to square centimeters instead of square feet
  mutate(BA_GROW = BA_GROW * 929.03,
         BA_GROW_AC = BA_GROW_AC * 929.03)
live
```

